The Forecast Trap

### 2022-02-28 manuscript/appendix-B.Rmd

\begin{equation}
f_i(Y_{t+1}) = Y_t + r_i Y_t \left( 1 - \frac{Y_t}{K_i} \right) * \xi_t(\sigma)
\end{equation}

where $\xi_i(t)$ is a multiplicative log-normal noise term with mean 1 and log-standard-deviation $\sigma_i$.

@@

for each prediction over 100 replicate simulations of 100 time steps each [Fig 1].

@@

````{r}
## Compute reward matrix (shared across all models)
n_s <- length(states)
n_a <- length(actions)
reward <- array(0, dim = c(n_s, n_a))
for (k in 1:n_s) {
  for (i in 1:n_a) {
    reward[k, i] <- reward_fn(states[k], actions[i])
  }
}
```

@@

# Simulations and step-ahead forecasts (Fig 2, 3A)

@@

Using the transition matrices directly, we can examine what each model would 
have forecast the future stock size to be in the following year when no fishing
occurs (note that for each model, we use the transition matrix that corresponds
to 'no fishing', `model[[state_index, ,1]]`) (Fig 1a, main text).

@@

# Proper scores (Fig 1)

@@

```{r proper_scores, message = FALSE}
# Gneiting & Raferty (2007), eq27
scoring_fn <- function(x, mu, sigma){ as.numeric( -(mu - x )^2 / sigma^2  - log(sigma) )}

@@

# Economic value tabulation (Fig 2B)

@@

Again we record the posterior belief in the 'model 1' parameter values, and stack the data frames for learning and planning-only cases into a single file used to generate figure 3B.

@@

For most of the simulation, the most probable $r$ value is around 0.3, and the most probable $K$ value around 1.5.

@@

To plot growth curves of individual models (Fig 4a), we evaluate

@@

The resulting model curves are shown in the main paper, Fig 4A.

>>>>

\begin{equation}
X_{t+1} = f_i(Y_t) = Y_t + r_i Y_t \left( 1 - \frac{Y_t}{K_i} \right) * \xi_t(\sigma)
\end{equation}

where harvest occurs before recruitment, $Y_t = \min(X_t - H_t, 0)$,  $\xi_i(t)$ is a multiplicative log-normal noise term with mean 1 and log-standard-deviation $\sigma_i$.

@@

for each prediction over 100 replicate simulations of 100 time steps each [Fig 2].

@@

```{r}
## Compute reward matrix (shared across all models)
n_s <- length(states)
n_a <- length(actions)
reward <- array(0, dim = c(n_s, n_a))
for (k in 1:n_s) {
  for (i in 1:n_a) {
    reward[k, i] <- reward_fn(states[k], actions[i])
  }
}
```

@@

# Simulations and step-ahead forecasts (Fig 3, 4A)

@@

Using the transition matrices directly, we can examine what each model would 
have forecast the future stock size to be in the following year when no fishing
occurs (note that for each model, we use the transition matrix that corresponds
to 'no fishing', `model[[state_index, ,1]]`) (Fig 2A, main text).

@@

# Proper scores (Fig 2C-D)

@@

```{r proper_scores, message = FALSE}
# Gneiting & Raferty (2007), eq27
scoring_fn <- function(x, mu, sigma){
  as.numeric( -(mu - x )^2 / sigma^2  - log(sigma) )
  }

@@

# Economic value tabulation (Fig 3B)

@@

Again we record the posterior belief in the 'model 1' parameter values, and stack the data frames for learning and planning-only cases into a single file used to generate figure 4B.

@@

For most of the simulation, the most probable $r$ value is around 0.3, and the most probable $K$ value around 10.

@@

To plot growth curves of individual models (Fig 5A), we evaluate

@@

The resulting model curves are shown in the main paper, Fig 5A.

### 2022-02-28 manuscript/round3-manuscript.Rmd

- abbreviated running title: "The Forecast Trap"
- Authorship statement: _This is a single-author paper_
- data accessibility statement: _All simulation data generated for analyses here, along with code required for the analysis, is available in the Zenodo Data archive, https://doi.org/10.5281/zenodo.4660621_
- number of words in the abstract: 148
- number of words in main text: 4900
- number of cited references: 33
- number of tables & figures: 5 Figures

@@

```{r, include=FALSE}
rmarkdown::render("appendix-A.Rmd")
rmarkdown::render("appendix-B.Rmd")
knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
                      warning = FALSE, dev="cairo_pdf", fig.height=5, fig.width=7.5)
```

@@

In this paper, I demonstrate that the model that makes the better prediction (defined as a strictly proper score, @Gneiting2007) is not necessarily the model that makes the better policy (defined in terms of utility, e.g. expected net present value, @Clark1990).
I show that our best methods for learning about model structure or parameters by repeatedly comparing forecasts to observations can be counter-productive.
Put another way, the value of information (VOI, as measured by the expected utility given that information minus the utility without it; see @Howard1966; @Katz1987), can actually be negative.
When VOI is negative, the decision-maker may become trapped into accepting mediocre outcomes derived from a model that makes accurate forecasts, even when a less accurate model that would generate better outcomes is available.
I will present two examples of this "forecast trap" and examine how it arises.

@@

To better understand how a model can produce a more accurate forecast and yet still lead to a worse decision, it may be helpful to start with a concrete example.
Figure \ref{fig1} shows both the forecasts and realized management outcomes of two alternative three-species models, "A" and "B" (see Appendix A) in predicting the population dynamics of striped bass (an economically important fishery) and double-crested cormorants (a target species for conservation) which both feed on a population of river herring (whose abundance we assume is not measured). 
Model A accurately forecasts the abundance of both bass and cormorants well into the future, but the optimal management strategy derived from its forecasts leads to steadily declining species abundances and overall disappointing net utility.
Model B produces substantially less accurate forecasts, but nevertheless achieves better outcomes.
The net utility under model B is in fact nearly identical to the maximum utility attainable given the true model, while management under model A achieves only 38% of that utility.

This management problem is motivated by a real world example of a herring fishery as described in @Brias2021, 
in which our manager seeks to balance multiple objectives of sustaining the cormorant population while maximizing the economic value from harvesting both herring and striped bass.
In the scenario depicted in Figure \ref{fig1}, I have used a the richer five-species model introduced @Brias2021 to drive the underlying dynamics, which includes three competing species of herring that are preyed upon by both the bass and the cormorants. 
Here, the manager seeks to maximize utility given by the weighted sum of the individual objectives [@Brias2021], placing 50% of the weight on the conservation objective and splits the remaining weight evenly over the harvests for predator (bass) and prey (herring) species.
I have assumed a partially observable system - in this case, the manager measures only the abundance of bass and cormorant species, and not of the three herring species.
In this scenario, I have further assumed the manager must choose a fixed fishing effort for herring and for the bass harvest,
I will consider more dynamic decision-processes later, but it is worth noting that in many real world conservation settings policy choices are highly constrained and frequent adjustment of those policies may be costly or impossible.
Likewise, the assumptions of partially observable system and imperfect models are characteristic of ecological decision-making.
Equations and code for all models in this example are presented in Appendix A.

@@

```{r figure4, fig.cap="Adaptive management under model uncertainty. Solid lines trace the trajectories of the state (fish stock, circles) and action (harvest quota, triangles), under adaptive management (learning). Dotted lines trace the corresponding trajectories if iterative learning is omitted, leaving the prior belief fixed throughout the simulation (planning). Color indicates the belief that model 1 is correct (blue), with an initial prior belief of 99%. Panel A: Management over the two candidate models, Model 1 and Model 2. Within a single iteration of adaptive management, the belief over models switches from a prior belief that heavily favored model 1 to a posterior that favors model 2 with near certainty. Future iterations reinforce the belief in model 2, resulting in both depressed harvests and low stock sizes (solid lines). If no iterative learning updates are performed, stock sizes and realized harvests (and thus economic profit) are both higher. Panel B: given 42 candidate models over a broad range of parameter values, adaptive management quickly reduces the probability of model 1, and substantially underperforms management without learning (dotted lines). While outcomes improve marginally relative to the two-model case (figure A) they remain significantly worse than had no iterative learning been included."}
am <- read_csv("../data/am.csv")
fig3a <- am %>% filter(time <=20) %>% 
  pivot_longer(c(stock, quota), values_to="stock", names_to="key") %>%
  mutate(key = factor(key, levels = c("stock", "quota"))) %>%
  ggplot(aes(time, stock, shape = key, group = interaction(method,key))) + 
  geom_line(aes(lty=method), col = "gray40", show.legend = FALSE) + 
  geom_point(aes(col = belief), size=2, show.legend = FALSE) +
  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) +
  ylab("fish stock") + labs(subtitle="A.")

@@

Third, managers may emphasize policy robustness over expected net utility. 
In many formal treatments, this is not qualitatively different to the analysis considered here: 
a manager simply chooses a different utility function, such as minimizing 'regret' rather than maximizing expected value [@Polasky2011]. 
Such approaches are just as vulnerable to bad outcomes (as defined by their own utility functions) whenever models are selected only on the basis of forecast skill.
But in practice, robust design may emphasize acceptable performance across the widest possible array of scenarios (e.g. candidate models).
This acts more like a sensitivity analysis of utility with respect to underlying assumptions, rather than an optimization routine.
Computationally, the former is much simpler, allowing researchers to evaluate the performance of a policy on more complex simulations for which calculating the optimal policy would be prohibitively difficult.

>>>>

- abbreviated running title: "The Forecast Trap"
- Authorship statement: _This is a single-author paper_
- data accessibility statement: _All simulation data generated for analyses here, along with code required for the analysis, is available in the Zenodo Data archive, https://doi.org/10.5281/zenodo.4660621_
- number of words in the abstract: 148
- number of words in main text: 4900
- number of cited references: 44
- number of tables & figures: 5 Figures

@@

```{r renderA, include=FALSE}
rmarkdown::render("appendix-A.Rmd")

```

```{r renderB, include=FALSE, eval=FALSE}
rmarkdown::render("appendix-B.Rmd")
```

```{r knit_global, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
                      warning = FALSE, dev="cairo_pdf", fig.height=5, fig.width=7.5)
```

@@

In this paper, I demonstrate that the model that makes the better prediction (defined as a strictly proper score, @Gneiting2007) is not necessarily the model that makes the better policy (defined in terms of utility, e.g. expected net present value, @Clark1990).
I show that our best methods for learning about model structure or parameters by repeatedly comparing forecasts to observations can be counter-productive.
Put another way, the value of information (VOI, as measured by the expected utility given that information minus the utility without it; see @Howard1966; @Katz1987), can actually be negative.
When VOI is negative, the decision-maker may become trapped into accepting mediocre outcomes derived from a model that makes accurate forecasts, even when a less accurate model that would generate better outcomes is available.
I will present two examples of this "forecast trap" and examine how it arises as a result of non-uniqueness of models [@Oreskes1994; @Schindler2015] with respect to either of these objectives.

@@

To better understand how a model can produce a more accurate forecast and yet still lead to a worse decision, it may be helpful to start with a concrete example.
Fig 1 shows both the forecasts and realized management outcomes of two alternative three-species models, "A" and "B" (see Appendix A) in predicting the population dynamics of striped bass (an economically important fishery) and double-crested cormorants (a target species for conservation) which both feed on a population of river herring (whose abundance we assume is not measured). 
Model A accurately forecasts the abundance of both bass and cormorants well into the future, but the optimal management strategy derived from its forecasts leads to steadily declining species abundances and overall disappointing net utility.
Model B produces substantially less accurate forecasts, but nevertheless achieves better outcomes.
The net utility under model B is in fact nearly identical to the maximum utility attainable given the true model, while management under model A achieves only 38% of that utility.

This management problem is motivated by a real world example of a herring fishery as described in @Brias2021, 
in which our manager seeks to balance multiple objectives of sustaining the cormorant population while maximizing the economic value from harvesting both herring and striped bass.
In the scenarios depicted in Fig 1, I have used a the richer five-species model introduced @Brias2021 to drive the underlying dynamics, which includes three competing species of herring that are preyed upon by both the bass and the cormorants. 
Here, the manager seeks to maximize utility given by the weighted sum of the individual objectives [@Brias2021], placing 50% of the weight on the conservation objective and splits the remaining weight evenly over the harvests for predator (bass) and prey (herring) species.
I have assumed a partially observable system - in this case, the manager measures only the abundance of bass and cormorant species, and not of the three herring species.
In this scenario, I have further assumed the manager must choose a fixed fishing effort for herring and for the bass harvest,
I will consider more dynamic decision-processes later, but it is worth noting that in many real world conservation settings policy choices are highly constrained and frequent adjustment of those policies may be costly or impossible.
Likewise, the assumptions of partially observable system and imperfect models are characteristic of ecological decision-making.
Equations and code for all models in this example are presented in Appendix A.

@@

```{r figure4, fig.cap="Adaptive management under model uncertainty. Solid lines trace the trajectories of the state (fish stock, circles) and action (harvest quota, triangles), under adaptive management (learning). Dotted lines trace the corresponding trajectories if iterative learning is omitted, leaving the prior belief fixed throughout the simulation (planning). Color indicates the belief that model 1 is correct (blue), with an initial prior belief of 99%. Panel A: Management over the two candidate models, Model 1 and Model 2. Within a single iteration of adaptive management, the belief over models switches from a prior belief that heavily favored model 1 to a posterior that favors model 2 with near certainty. Future iterations reinforce the belief in model 2, resulting in both depressed harvests and low stock sizes (solid lines). If no iterative learning updates are performed, stock sizes and realized harvests (and thus economic profit) are both higher. Panel B: given 42 candidate models over a broad range of parameter values, adaptive management quickly reduces the probability of model 1, and substantially underperforms management without learning (dotted lines). While outcomes improve marginally relative to the two-model case (panel A) they remain significantly worse than had no iterative learning been included."}
am <- read_csv("../data/am.csv")
fig3a <- am %>% filter(time <=20) %>% 
  pivot_longer(c(stock, quota), values_to="stock", names_to="key") %>%
  mutate(key = factor(key, levels = c("stock", "quota"))) %>%
  ggplot(aes(time, stock, shape = key, group = interaction(method,key))) + 
  geom_line(aes(lty=method), col = "gray40", show.legend = FALSE) + 
  geom_point(aes(col = belief), size=2, show.legend = FALSE) +
  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) +
  ylab("fish stock") + labs(subtitle="A.")

@@

Third, managers may emphasize policy robustness over forecast skill [@Schindler2015]. 
In many formal treatments, this is not qualitatively different to the analysis considered here: 
a manager simply chooses a different utility function, such as minimizing 'regret' rather than maximizing expected value [@Polasky2011]. 
Such approaches are just as vulnerable to bad outcomes (as defined by their own utility functions) whenever models are selected only on the basis of forecast skill.
But in practice, robust design may emphasize acceptable performance across the widest possible array of scenarios (e.g. candidate models).
This acts more like a sensitivity analysis of utility with respect to underlying assumptions, rather than an optimization routine [e.g. @Punt2016; @Fischer2009].
Computationally, the former is much simpler, allowing researchers to evaluate the performance of a policy on more complex simulations for which calculating the optimal policy would be prohibitively difficult.
